---
title: "Final Project 4a: Final Dashboard and GitHub Pages Deployment"
author: "Ramya Rajendran"
date: "`r Sys.Date()`"
output:
  flexdashboard::flex_dashboard:           # Use flexdashboard layout
    orientation: rows                      # Stack content in horizontal rows
    vertical_layout: fill                  # Fill vertical space for responsiveness
    theme: flatly                          # Apply the 'flatly' Bootstrap theme
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r libraries}
library(flexdashboard)
library(plotly)
library(DT)
library(dplyr)
library(readr)
```

Introduction
=======================================================================

Row {.tabset .tabset-fade}
-----------------------------------------------------------------------


### 1. Research Questions & Motivation

**Project Overview**

This interactive dashboard provides a comprehensive **exploratory analysis** of **Early-Onset Colorectal Cancer (EOCRC)** using data from the **Surveillance, Epidemiology, and End Results (SEER) program**. The dataset includes **demographic, clinical, and socioeconomic variables** linked to colorectal cancer diagnoses between **2018 and 2021**. The goal is to enable data-driven decision-making for researchers and policymakers.

This study investigates disparities in **early-onset colorectal cancer (EOCRC)** and **late-onset colorectal cancer (LOCRC)**, focusing on demographic, socioeconomic, and geographic factors.

**Key questions:**

- How do **age, sex, and race/ethnicity** influence stage at diagnosis?
- What is the relationship between **household income** and stage of EOCRC?
- How do **geographic disparities** impact access and diagnosis?
- What **clinical factors** affect EOCRC development in adults aged **18–49**?


### 2. Objectives
- **Analyze demographic patterns** in EOCRC cases.
- **Explore disparities** across race, sex, and socioeconomic status.
- **Investigate urban vs. rural classification** and its impact on cancer diagnosis.
- **Enable interactive data filtering** to facilitate real-time analysis and insights.


### 3. Data Source

- The data source is the SEER Program (https://seer.cancer.gov/), a trusted U.S. cancer surveillance system that requires an account to access the data. There were 144,788 cases (EOCRC: ages 18-49, LOCRC: ages 50+). The data were collected using hospital-based cancer registries, pathology reports, and medical records. The study population consisted of diverse EOCRC and LOCRC patients across multiple geographic regions in the U.S. The time period of the data was 2018-2021, capturing recent trends in cancer disparities.


Key Variables for Analysis
=======================================================================

Row {.tabset .tabset-fade}
-----------------------------------------------------------------------

### 1. Demographics

- **Age Groups:** **18-49** (EOCRC), **50+** (LOCRC)

- **Sex:** Male, Female

- **Race/Ethnicity:** 
  - Non-Hispanic White
  - Non-Hispanic Black
  - Hispanic
  - Non-Hispanic Asian/Pacific Islander
  - Non-Hispanic American Indian/Alaska Native

### 2. Cancer Details
- **Tumor Site:** Location in the colon
- **Histology:** Type of cancer cells
- **Tumor Behavior:** Benign, in situ, malignant
- **Stage at Diagnosis:** Localized, regional, distant

### 3. Socioeconomic Factors
- **Household Income:** Inflation-adjusted to **2022** (categorized)
- **Rural vs. Urban:** Metropolitan or nonmetropolitan areas

### 4. Diagnosis & Clinical Data
- **Year of Diagnosis:** **2018-2021**
- **Diagnostic Method:** Histology, cytology, imaging, clinical
- **Reporting Source:** Hospital, private lab, physician’s office

### 5. Genetic & Tumor Markers
- **KRAS Mutation:** Present or absent
- **Tumor Grade:** Low, intermediate, or high


Planned Interactive HTML Widgets
=======================================================================

Row {.tabset .tabset-fade}
-----------------------------------------------------------------------

### 1. Interactive Bar Chart (Plotly - EOCRC Cases by Race/Ethnicity)

**Description:**

- **X-Axis:** Race/Ethnicity categories

- **Y-Axis:** Number of EOCRC/LOCRC cases

- **Format:** Stacked bar chart

### 2. Interactive Features

- **Hover Tooltips:** Displays exact case counts for each racial/ethnic group.
- **Legend Toggle:** Allows users to filter selected racial/ethnic groups.
- **Zooming & Panning:** Enables in-depth analysis of disparities.

### 3. Purpose

- Identify racial and ethnic disparities in colorectal cancer incidence.
- Analyze potential risk factors for different populations.
- Assess **healthcare access inequalities** affecting early detection.


EOCRC Visualization Bar Chart
=======================================================================

Row
-----------------------------------------------------------------------

Column {data-width=90% data-height=4000}
-----------------------------------------------------------------------

### Interactive Bar Chart (Plotly - EOCRC Cases by Race/Ethnicity)

```{r load_data_bar}
# Load and preprocess dataset:
# - Ensure all columns are read as character to prevent unwanted type conversions
# - Convert race/ethnicity column to a factor with meaningful labels
# - Clean and convert age column to numeric by removing non-numeric characters
# - Create a binary variable to classify early-onset cases (ages 18-49) vs. late-onset

# Read dataset with explicit column types to prevent automatic conversions

#data <- read_csv("550_thesis_dataset_number_format.csv", col_types = cols(.default = "c"), n_max = 1000)
data <- read_csv("550_thesis_dataset_number_format.csv",
                 col_types = cols(.default = "c"))

# Convert categorical variables to factors for proper filtering and analysis
data$race_ethnicity <- factor(
  data$`Race and origin recode (NHW, NHB, NHAIAN, NHAPI, Hispanic)`,
  levels = c(1, 2, 3, 4, 5),
  labels = c(
    "Non -Hispanic White",
    "Non-Hispanic Black",
    "Non-Hispanic American Indian/Alaska Native",
    "Non-Hispanic Asian/Pacific Islander",
    "Hispanic"
  )
)

# Convert age to numeric by removing any non-numeric characters (e.g., text entries)
data$`Age recode with single ages and 90+` <- as.integer(gsub("[^0-9]", "", data$`Age recode with single ages and 90+`))

# Create early_onset_yes_or_no variable (Binary classification: 1 = Early-Onset, 0 = Late-Onset)
data$early_onset_yes_or_no <- ifelse(
  data$`Age recode with single ages and 90+` >= 18 &
    data$`Age recode with single ages and 90+` <= 49,
  1,
  0
)
```


```{r plotly_bar_chart}
# Aggregate data to count EOCRC cases by race/ethnicity and onset category
# Define a bright color palette for better visual distinction
# Create an interactive stacked bar chart using Plotly:
#   - X-axis: Racial/ethnic groups
#   - Y-axis: Number of EOCRC cases
#   - Color: Differentiates early-onset (18-49) and late-onset (50+) cases
#   - Enhanced visualization with clear labels, legend, and spacing adjustments

race_summary <- data %>%
  group_by(race_ethnicity, early_onset_yes_or_no) %>%
  summarise(total_cases = n(), .groups = "drop")

# Define bright color palette for improved visual appeal
color_palette <- c("18-49" = "Salmon", "50+" = "Sky Blue")

# Create interactive bar chart using Plotly with enhanced color contrast
plotly_bar <- plot_ly(
  data = race_summary,
  x = ~ race_ethnicity,
  # X-axis represents racial/ethnic groups
  y = ~ total_cases,
  # Y-axis represents the count of EOCRC cases
  color = ~ factor(early_onset_yes_or_no, labels = c("50+", "18-49")),
  # Color coding for age groups
  colors = color_palette,
  type = 'bar',
  marker = list(line = list(width = 1.5))
) %>%
  layout(
    title = list(text = "EOCRC Cases by Race/Ethnicity and Age Group", y = 0.95),
    xaxis = list(title = "Race/Ethnicity"),
    yaxis = list(title = "Number of Cases", tickformat = "d"),
    margin = list(t = 80),
    # Add space above the title
    barmode = "stack",
    # Stack bars for better visualization
    legend = list(title = list(text = "Age Group"))
  )

plotly_bar
```

Column {data-width=10%}
-----------------------------------------------------------------------

### Key takeaways description

Key takeaways for bar graph: The majority of colorectal cancer cases occurred in people aged 50+. Non-Hispanic Whites made up more than 50% of all CRC cases, while Non-Hispanic American Indians/Alaska Natives made up less than 1%  of all CRC cases.

Data Table (DT) Description
=======================================================================

Row {.tabset .tabset-fade}
-----------------------------------------------------------------------

### 1. Filterable Data Table (DT) – DescriptionPatient Demographics & Clinical Characteristics

**Description:**

**searchable & sortable** table for patient demographics and clinical data.

- **X-Axis:** Race/Ethnicity categories
- **Y-Axis:** Number of EOCRC/LOCRC cases
- **Format:** Stacked bar chart

### 2. Interactive Features
- **Dropdown Filters:** Age, sex, race/ethnicity, tumor stage, income level.
- **Search Functionality:** Quick lookup of patient records.
- **Sortable Columns:** Diagnose trends over time.
- **Pagination & Adjustable Display:** Customizable data views.

### 3. Purpose
- Compare **EOCRC vs. LOCRC** patient characteristics.
- Identify patterns in cancer progression.
- Support **research & policy interventions** for cancer prevention.


Filterable Data Table (DT)
=======================================================================

Row
-----------------------------------------------------------------------

Column {data-width=90% data-height=4000}
-----------------------------------------------------------------------

### Filterable Data Table (DT) – Patient Demographics & Clinical Characteristics


```{r load_data_dt}
# Load dataset with explicit column types to prevent automatic conversions (limit to 1000 rows)
# Convert necessary columns to appropriate data types for better filtering and visualization:
#   - Patient ID: Integer for numerical processing
#   - Categorical variables: Converted to factors for analysis
#   - Age: Cleaned and converted to integer by removing non-numeric characters
#   - Year of diagnosis: Converted to integer
#   - Behavior code: Kept as character for searchability in DT tables
#   - Create 'early_onset_yes_or_no' as a binary factor (Yes = 18-49, No = 50+)



#data <- read_csv("550_thesis_dataset_full_format.csv", col_types = cols(.default = "c"), n_max=1000)
data <- read_csv("550_thesis_dataset_full_format.csv",
                 col_types = cols(.default = "c"))

# Convert necessary columns to correct data types for better filtering and visualization
data <- data %>%
  mutate(
    `Patient ID` = as.integer(`Patient ID`),
    # Convert ID to integer for numerical processing
    `Sex` = as.factor(`Sex`),
    # Convert categorical variables to factors
    `Combined Summary Stage (2004+)` = as.factor(`Combined Summary Stage (2004+)`),
    `Rural-Urban Continuum Code` = as.factor(`Rural-Urban Continuum Code`),
    `Median household income inflation adj to 2022` = as.factor(`Median household income inflation adj to 2022`),
    `Marital status at diagnosis` = as.factor(`Marital status at diagnosis`),
    `SEER registry (with CA and GA as whole states)` = as.factor(`SEER registry (with CA and GA as whole states)`),
    `Age recode with single ages and 90+` = as.integer(gsub(
      "[^0-9]", "", `Age recode with single ages and 90+`
    )),
    # Remove non-numeric characters
    `Year of diagnosis` = as.integer(`Year of diagnosis`),
    `Race and origin recode (NHW, NHB, NHAIAN, NHAPI, Hispanic)` = as.factor(`Race and origin recode (NHW, NHB, NHAIAN, NHAPI, Hispanic)`),
    `KRAS (2018+)` = as.factor(`KRAS (2018+)`),
    `Diagnostic Confirmation` = as.factor(`Diagnostic Confirmation`),
    `Derived Summary Grade 2018 (2018+)` = as.factor(`Derived Summary Grade 2018 (2018+)`),
    `Behavior code ICD-O-3` = as.character(`Behavior code ICD-O-3`),
    # Ensure proper searchability in DT tables
    `Type of Reporting Source` = as.factor(`Type of Reporting Source`),
    `early_onset_yes_or_no` = as.factor(
      ifelse(
        `Age recode with single ages and 90+` >= 18 &
          `Age recode with single ages and 90+` <= 49,
        "Yes",
        "No"
      )
    )
  )
```

 
```{r create_table}
# Prepare and format data table for dashboard usability:
#   - Select relevant columns and rename them for clarity
#   - Ensure categorical and numerical fields are properly labeled
#   - Create an interactive data table using DT:
#       - Display 10 rows per page by default
#       - Enable automatic column width adjustment
#       - Highlight search terms for better usability
#       - Allow column-wise filtering for precise data exploration

# Unwanted variables which are not used in thesis are commented out below.
data_table <- data %>%
  select(
    #`Patient ID` = `Patient ID`,
    #`Sequence Number` = `Sequence number`,
    `Sex` = `Sex`,
    `Summary Stage (2004+)` = `Combined Summary Stage (2004+)`,
    `Rural-Urban Classification` = `Rural-Urban Continuum Code`,
    `Median Household Income (2022 Adjusted in $)` = `Median household income inflation adj to 2022`,
    #`Marital Status at Diagnosis` = `Marital status at diagnosis`,
    `SEER Registry (CA & GA as Whole States)` = `SEER registry (with CA and GA as whole states)`,
    `Age at Diagnosis` = `Age recode with single ages and 90+`,
    `Year of Diagnosis` = `Year of diagnosis`,
    `Race & Ethnicity` = `Race and origin recode (NHW, NHB, NHAIAN, NHAPI, Hispanic)`,
    # `KRAS Mutation (2018+)` = `KRAS (2018+)`,
    # `Diagnostic Confirmation` = `Diagnostic Confirmation`,
    # `Tumor Grade (2018+)` = `Derived Summary Grade 2018 (2018+)`,
    # `Histologic Type (ICD-O-3)` = `Histologic Type ICD-O-3`,
    `Histology (Broad Grouping)` = `Histology recode - broad groupings`,
    # `Behavior Code (ICD-O-3)` = `Behavior code ICD-O-3`,
    # `Type of Reporting Source` = `Type of Reporting Source`,
    `Early-Onset Cancer (18-49) (Yes/No)` = `early_onset_yes_or_no`
  )
datatable(
  data_table,
  options = list(
    pageLength = 10,
    # Default number of rows per page
    autoWidth = TRUE,
    # Adjust column widths automatically
    searchHighlight = TRUE,
    # Enable search term highlighting
    dom = 'lrtip'  # Configure table layout (search box, processing, pagination)
  ),
  rownames = FALSE,
  filter = "top"
)  # Enable column-wise filtering
```


Column {data-width=10%}
-----------------------------------------------------------------------

### Key takeaways description

Key takeaways from filterable data table: Approximately 47% of all colorectal cancer cases were from females and 53% were from males. 126,003 cases were LOCRC, while about 18,785 cases were EOCRC. 2020 had the lowest number of CRC cases with 33,582 cases.




Summary
=======================================================================

Row {.tabset .tabset-fade}
-----------------------------------------------------------------------

### 1. Conclusion

This dashboard facilitates **interactive exploration** of **Early-Onset Colorectal Cancer (EOCRC)**, providing:

- **Dynamic filtering** of key clinical and demographic data.
- **Visualization of disparities** in cancer incidence.
- **Insights for researchers & policymakers** in understanding EOCRC trends.

### 2. Future Enhancements
- **Integration of machine learning models** for predictive analysis.
- **Geo-spatial mapping** to explore regional cancer disparities.
- **Advanced statistical modeling** to analyze survival trends.

This tool serves as a foundation for further **data-driven public health research** and **clinical interventions**


Link to GitHub Repository
=======================================================================

Row
-----------------------------------------------------------------------

🔗 Access the full GitHub repository here, which includes the project’s source code, README.md, rendered HTML dashboard, and all associated R Markdown (.Rmd) files:

👉 https://github.com/ramya-rajendran-at-emory-edu/eocrc_and_locrc_dashboard/

